Northern Pikeminnow Predation on Juvenile Chinook Salmon in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Populations

### 2021-12-28 analysis/paper/paper.Rmd
edits to m-r and gastric discussion

<!--
Re-hash out our assumptions? What assumptions were most likely violated e.g., the closed population assumption? If we assumed an open population with varying immigration/emigration rates how might that affect our estimate? In the end, how good do we think our estimate is and is the bottom-line that we belive there still to be a shit-ton of pikeminnow even if our assumptions were violated?
-->

All estimators suggest a large number of Northern Pikeminnow occupying Deadwater Slough. <!-- This might be a good place to bring in some comparisons with Northern Pikeminnow densities from elsewhere to put our numbers in context.-->

If the first mark-recapture assumption was violated and the population is actually open (potential immigration and emigration during the survey period), as long as the immigration and emigration rates are equal between marked and unmarked fsih, our estimators should still provide unbiased estimates of abundance. Given the size of the sampling area, and the length of the sampling period (two weeks).

We are comfortable with assumptions 4 and 5 (no loss of marks and no mistaken or missed marks). Assumptions 2 and 3 indicate whether the second (and subsequent) surveys are random samples of the population. Angling methods may provide a size selection bias, but that may only affect the interpretation of the abundance estimates (i.e., our estimate are the abundance of Northern Pikeminnow above some size threshold). Unequal catchability between individual fish is very difficult to assess, especially if that heterogeneity arises from being caught once.

<!-- If catching and marking fish makes them want to leve the Deadwater area (more than unmarked fish), that could be making our estimates biased high (due to fewer recaptures). I don't imagine this is happening, but does anyone else have concerns about that? -->

The spring abundance estimate is smaller than either fall sampling event. However, our approach assumes equal capture probabilities betwee the fall and spring sampling events. If it does differ, we believe that the capture probability in the spring could be lower due to higher flows, which would lead to our spring abundance estimate being biased low. Therefore, we presume our abundance estimate of Northern Pikeminnow for the spring 2021 sampling event to be a conservative estimate.

@@

The majority of fish collected during this study received gastric lavage, including some non-predatory species. To validate the efficacy of this method, we euthanized nine Northern Pikeminnow after gastric lavage was completed and removed the remaining stomach contents via dissection. Our results support previous findings that gastric lavage effectively removes stomach contents (@Lott2020). Additionally, we found that the fish captured in net-traps had a similar proportion of stomach content samples. Of those, 1,214 (76.2%) were completely empty, 345 (21.6%) had stomach contents, and 35 (2.2%) contained fish remnants. We were successfully able to identify Redside Shiners, a Largescale Sucker, sculpin, Mountain Whitefish, and one juvenile Chinook Salmon in the stomach contents. However, decomposition from digestion rendered most stomach contents unidentifiable.

>>>>

All mark-recapture abundance estimators suggest a large number of Northern Pikeminnow occupying Deadwater Slough. Estimators assumed: (1) the population is closed (no immigration, emigration, births, or deaths during the sampling period), (2) all fish have equal chance of being caught in the second (and subsequent) samples(s), (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, and (5) marks are not missed or mistaken. If the population is actually open (potential immigration and/or emigration during the survey period), our estimators should still provide unbiased estimates of abundance as long as the immigration and emigration rates are equal between marked and unmarked fish. Given the size of the sampling area, and the limited length of the sampling period (two weeks), we are comfortable with the assumption of no immigration or emigration. Assumption (3) indicates whether the second (and subsequent) surveys are random samples of the population. Angling methods may provide a size selection bias, but that only affects the interpretation of the abundance estimates (i.e., our estimates are the abundance of Northern Pikeminnow susceptible to angling above some size threshold). Unequal catchability between individual fish is very difficult to assess, especially if that heterogeneity arises from being caught once; however, we did recapture two individual fish more than once suggesting that marked fish were still susceptible to angling. We are also comfortable with assumptions 4 and 5 (no loss of marks and no mistaken or missed marks) given that marks were physical (hole punches in fins) and that each fish was examined on a measuring board to collect lengths.

We do consider two potential violations of assumption, both which reduces the probability that a marked fish is susceptible to capture in the second (and subsequent) samples. First, the assumption of a closed population could be violated if some marked fish that are released back to the population die. We don't believe this case to be likely, as fish that were dead or nearly so after processing were not released back to the population (nor recorded as such), and those that were released were hand-released only if they were able to swim away on their own volition; however, the potential for marked and released fish to have died at a higher rate than unmarked fish should be acknowledged. The second, and more likely, reason is that marked and released fish did not mix thoroughly back into the population (at least not at a fast enough rate), reducing their probability of capture in subsequent captures, thus violating assumption (2). We attempted to release marked fish throughout Deadwater Slough to facilitate mixing back into the population; however, anecdotally volunteer anglers would congregate to particular areas of the slough either to socialize with other anglers, because the fishing was good, etc. and if marked and released fish were not readily available in those areas that would make marked fish less susceptible to capture. This violation would, in effect, reduce the number of marked fish available for recapture or reduce the marking fraction in some areas of the slough. Violations of assumption (2) would lead to an overestimate of abundance of Northern Pikeminnow which is a real possibility in our study.

The spring abundance estimate was smaller than either fall abundance estimate. However, our approach assumed equal capture probabilities between the fall and spring sample efforts. If it does differ, we suspect that the capture probability in the spring would be lower due to higher flows (i.e., reduced densities), which would lead to our spring abundance estimate being biased low. Therefore, we presume our abundance estimate of Northern Pikeminnow for the spring 2021 effort to be a conservative estimate.

@@

The majority of fish caught during the study received gastric lavage. To evaluate the efficacy of the method, we euthanized nine Northern Pikeminnow after gastric lavage was completed and removed the remaining stomach contents via dissection. In each case, no additional stomach contents were identified, suggesting that gastric lavage was effective in removing stomach contents, although our sample size was low. However, in total, we only identified fish or fish parts in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r round(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"] * 100, 1)`%) of the `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow examined using gastric lavage. Aside from juvenile Chinook Salmon, identified species included Redshine Shiner, Largescale Sucker *Catostomus macrocheilus*, sculpin *Cottus* spp., and Mountain Whitefish *Prosopium williamsoni*; however, decomposition from digestion rendered most stomach contents unidentifiable.

Although we only identified fish or fish parts within `r round(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"] * 100, 1)`% of Northern Pikeminnow examined and that fish remnants were only `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of the overall stomach contents, we believe those numbers to be misleading.
